The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-05-23 ms.Rmd
switching to spThin for module example, combining figures for two Feng workflows into one multi-panel figure

@@

![Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`. White areas are masked due to being in the sea or outside the MESS mask. \label{fig:fengmap}](./figs/fengworkflow-1.png)

The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.

Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and  predictions on a zoomable map, and to inspect the response curves of the fitted model. These outputs are shown in Figure \ref{fig:interactive}.

```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide"}
Feng_Papes_Update <- 
  ChangeWorkflow(workflow = Feng_Papes,
                   output = Chain(InteractiveMap, 
                                  ResponseCurve(cov = 1)))
```

```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 300, fig.show = "hide", fig.height = 3.2}
# combine the printed map and first effect plot

# load both images
r_map <- brick('figs/interactive_map.png')
r_resp <- raster('figs/fengChangeWorkflow-2.png')
r_resp <- brick(r_resp, r_resp, r_resp)

# set up layout so that heights are the same for both panels
plot_height <- min(nrow(r_map), nrow(r_resp))

# get rescaling factor for each
rescale_map <- plot_height / nrow(r_map)
rescale_resp <- plot_height / nrow(r_resp)

# get widths
width_map <- ncol(r_map) * rescale_map
width_resp <- ncol(r_resp) * rescale_resp

# set up layout (maximum total of 200 columns) with a gap in between
# mar won't work for these
gap <- 2
widths <- c(width_map, width_resp)
widths <- round(widths * ((200-gap) / sum(widths)))
widths <- c(widths[1], gap, widths[2])
mat <- matrix(rep(1:3, widths),
              nrow = 1)
layout(mat)

@@

@@

![The outputs of the updated workflow object `Feng_Papes_Update`. A) A screenshot of the interactive map produced by the `InteractiveMap`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results. B) A response curve produced by the `ResponseCurve` for the first covariate, bio1. \label{fig:interactive}](./figs/combinefengupdate-1.png)

@@

#### Example 2. Building a point process model module

Recent SDM research has identified an equivalence between MaxEnt and Poisson point process models (PPMs) [] and noted that PPMs can be fitted using Poisson modelling software [Warton & Renner, Renner et al.]. This means that other modelling approaches can be fitted as PPMs, after a bit of data preparation. We illustrate how to build a `zoon` model module by defining a module to fit a PPM using a Generalized Additive Model (GAM).

The code for our model is as follows (see Renner et al., 2014 for details of this approach):

```{r bioclimmodule, eval = FALSE}
PPM_GAM <- function () {
    # to do...
  }
```

The only complicated aspect of this code is construction of a `zoonModel` object, which contains code to make predictions from the model in a standardised way. The other module types do not require this step, so are simpler to implement. The code also meets the minimal requirements of a *model* module, accepting the required argument `.df` and returning the `zoonModel` object.

To convert this code into a ZOON module, we need to write it to a standalone file with the required metadata. the `zoon` function `BuildModule` helps with this step, and can also run checks to make sure we got everything right:

```{r buildmodule, eval = FALSE} 
BuildModule(PPM_GAM, ...)
```

This module can now be used in a `zoon` workflow and shared with others. To upload the module to the `zoon` modules repository, [TBD].

>>>>

##### Accessing module documentation

<!-- to do -->

@@

This module plots a static map of the predicted distribution, and the underlying data.
This map is shown in \ref{fig:feng_papes_plots}A.
The resulting workflow contains all the code required to re-run the workflow, the input data and the results of executing each module.
The object `Feng_Papes` could therefore be saved as a binary file and shared as a reproducible representation of this research.

Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model. These outputs are shown in Figure This map is shown in \ref{fig:feng_papes_plots}, panels B and C.

```{r fengChangeWorkflow, eval = TRUE, fig.show = "hide", fig.height = 9}
Feng_Papes_Update <- 
  ChangeWorkflow(workflow = Feng_Papes,
                   output = Chain(InteractiveMap, 
                                  ResponseCurve(cov = 1)))
```

```{r combinefengupdate, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 3.2}
# combine the static and interactive maps and first effect plot

# load both images
r_map1 <- brick('figs/fengworkflow-1.png')
r_map <- brick('figs/interactive_map.png')
r_resp <- raster('figs/fengChangeWorkflow-1.png')
r_resp <- brick(r_resp, r_resp, r_resp)

# set up layout so that heights are the same for both panels
plot_height <- min(nrow(r_map1), nrow(r_map), nrow(r_resp))

# get rescaling factor for each
rescale_map1 <- plot_height / nrow(r_map1)
rescale_map <- plot_height / nrow(r_map)
rescale_resp <- plot_height / nrow(r_resp)

# get widths
width_map1 <- ncol(r_map1) * rescale_map1
width_map <- ncol(r_map) * rescale_map
width_resp <- ncol(r_resp) * rescale_resp

# set up layout (maximum total of 200 columns) with a gap in between
# mar won't work for these
gap <- 2
widths <- c(width_map1, width_map, width_resp)
widths <- round(widths * ((200 - (2 * gap)) / sum(widths)))
widths <- c(widths[1], gap, widths[2], gap, widths[3])
mat <- matrix(rep(1:5, widths),
              nrow = 1)
layout(mat)

# plot statis map
plotRGB(r_map1, maxpixels = Inf)

@@

# add a panel letter
mtext(text = 'C',
        side = 3,
        line = -3,
        adj = 0)

@@

![Outputs of the workflow objects `Feng_Papes` and `Feng_Papes_Update`. A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap` module in the workflow `Feng_Papes`. White areas are masked due to being in the sea or outside the MESS mask. B) A screenshot of the interactive map produced by the `InteractiveMap` moduels in the workflow `Feng_Papes_Update`, displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results. C) A response curve produced by the `ResponseCurve` module for the first covariate, bio1 in the workflow `Feng_Papes_Update`. \label{fig:feng_papes_plots}](./figs/combinefengupdate-1.png)

@@

#### Example 2. Building a spatial thinning module

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the `zoon` data into the spThin format, uses this package to apply the algorithm, and converts data back into `zoon`'s expected format again:

```{r spthinmodule, eval = TRUE}
spThin <- function (.data, thin = 50) {

  # check these are presence-background data
  stopifnot(all(.data$df$type %in% c('presence', 'background')))

  # install & load the package
  zoon::GetPackage('spThin')

  # get dataframe & index to presence data
  df <- .data$df
  pres_idx <- which(df$type == 'presence')

  # prepare presence data subset and apply thinning
  sub_df <- data.frame(LAT = df$latitude[pres_idx],
                       LONG = df$longitude[pres_idx],
                       SPEC = NA)
  th <- thin(loc.data = sub_df,
             thin.par = thin,
             reps = 1,
             locs.thinned.list.return = TRUE,
             write.files = FALSE,
             write.log.file = FALSE)

  # get index to rows in sub_df, update the full dataset and return
  pres_keep_idx <- rownames(th[[1]])
  .data$df <- rbind(df[pres_idx[pres_keep_idx], ],
                    df[-pres_idx, ])
  return (.data)
}
```

To convert this code into a ZOON module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:

```{r buildmodule, eval = TRUE, results = 'hide'} 
BuildModule(object = spThin,
            type = 'process',
            title = 'Spatial thinning of Presence-only Data',
            description = paste('Apply the stochastic spatial thinning',
                                'algorithm implemented in the spThin',
                                'package to presence data in a',
                                'presence-background dataset'),
            details = paste('Full details of the algorithm are available in',
                            'the open-access article by Aiello-Lammens',
                            'et al. (2015): dx.doi.org/10.1111/ecog.01132'),
            author = 'ZOON Developers',
            email = 'zoonproject@gmail.com',
            paras = list(thin = paste('Thinning parameter - the required',
                         'minimum distance (in kilometres) between points',
                         'after applying the thinning procedure')),
            dataType = 'presence-only',
            check = TRUE)
```

This module can now be shared so that others can use it in their ZOON workflows. The bexst way of sharing it is via the `zoon` modules repository. Modules can uploaded there, via the module submission system at [URL TBD].
